Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-12-24 ms/manuscript.Rmd
Revise intro

To better understand the role of the canopy in fern diversification, one obvious direction is to explore putatively adaptive fern epiphyte traits in a phylogenetic context. 
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012]. 
Previous studies have identified multiple traits as likely adaptations to these conditions, albeit many without consideration of phylogeny. 
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a phylogenetic context including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* @Schuettpelz2009.

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities and reduced phylogenetic signal of relevant traits. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

>>>>

To better understand the role of the canopy in fern diversification, one obvious direction is to explore putatively adaptive fern epiphyte traits in a phylogenetic context [@Givnish2015]. 
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature [@Zotz2001; @WatkinsJr.2009; @Lowman2012]. 
Previous studies have identified multiple traits as likely adaptations to these conditions, albeit many without consideration of phylogeny [\eg, @Kluge2007]. 
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a broad phylogenetic context including both life stages.

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kromer2005; @Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.
